---
title: "Data Preparation for JustReinvest BOCSAR Dashboard - with censoring"
author: "Owen Forbes"
format: html
editor: source
---


## Final step before joining with spatial LGA shapefile = CENSOR cell counts <5 with value 'x' - keep in mind for processing/vis that x's are present for censored cells

## To deploy shiny app:
rsconnect::deployApp("/Users/Current/Desktop/Good_Data_Institute/JustReinvest/code/GDI_BOCSAR_Dashboard/code/dashboard_shiny_v3", account = "ofor22", server = "shinyapps.io")

# Load required libraries
```{r}
library(tidyverse)
library(readxl)
library(sf)
library(rmapshaper)

setwd("~/Desktop/Good_Data_Institute/JustReinvest/code/GDI_BOCSAR_Dashboard/code")
```


# Load required libraries
```{r}
library(tidyverse)
library(readxl)
library(sf)
library(rmapshaper)
```

# Define LGA name mapping and standardization function
```{r}
lga_name_mapping <- c(
  "Bathurst Regional" = "Bathurst",
  "Central Coast" = "Central Coast (NSW)",
  "Coffs Harbour" = "Coffs Harbour",
  "Dubbo Regional" = "Dubbo",
  "Mid-Western Regional" = "Mid-Western",
  "Queanbeyan-Palerang Regional" = "Queanbeyan-Palerang",
  "Snowy Monaro Regional" = "Snowy Monaro",
  "Upper Hunter Shire" = "Upper Hunter",
  "Port Macquarie-Hastings" = "Port Macquarie-Hastings",
  "Glen Innes Severn" = "Glen Innes Severn",
  "Liverpool Plains" = "Liverpool Plains",
  "Tamworth Regional" = "Tamworth",
  "Byron" = "Byron",
  "Richmond Valley" = "Richmond Valley",
  "Cootamundra-Gundagai" = "Cootamundra-Gundagai",
  "Snowy Valleys" = "Snowy Valleys",
  "The Hills Shire" = "The Hills",
  "Bayside" = "Bayside (NSW)",
  "Inner West" = "Inner West",
  "Northern Beaches" = "Northern Beaches",
  "Campbelltown" = "Campbelltown (NSW)",
  "Sutherland Shire" = "Sutherland",
  "Upper Lachlan Shire" = "Upper Lachlan",
  "Greater Hume Shire" = "Greater Hume",
  "Warrumbungle Shire" = "Warrumbungle",
  "Armidale Regional" = "Armidale",
  "Ku-Ring-Gai" = "Ku-ring-gai"
)

standardize_lga_name <- function(name) {
  name <- sub(" \\(NSW\\)$", "", name)
  if (name %in% names(lga_name_mapping)) {
    return(lga_name_mapping[name])
  }
  return(name)
}
```

# Read and process BOCSAR data including POI
```{r}
bocsar_path <- "/Users/Current/Desktop/Good_Data_Institute/JustReinvest/code/GDI_BOCSAR_Dashboard/data/BOCSAR/BOCSAR_New/sw24-23906 Contacts with the CJS by Aboriginality and LGA_V1.xlsx"
poi_path <- "/Users/Current/Desktop/Good_Data_Institute/JustReinvest/code/GDI_BOCSAR_Dashboard/data/BOCSAR/BOCSAR_New/sr24-23842_BOCSAR_PersonsOfInterest.xlsx"

clean_dataframe <- function(df) {
  df <- df[rowSums(is.na(df)) != ncol(df), ]
  df <- df[!is.na(df[[1]]), ]
  names(df) <- make.names(names(df), unique = TRUE)
  return(df)
}

bocsar_sheets <- excel_sheets(bocsar_path)[1:8]
bocsar_data <- list()

for (sheet in bocsar_sheets) {
  sheet_data <- read_excel(bocsar_path, sheet = sheet, skip = 5, col_names = TRUE)
  sheet_data <- clean_dataframe(sheet_data)
  bocsar_data[[sheet]] <- sheet_data
}

poi_data <- read_excel(poi_path, skip = 4, col_names = TRUE)
poi_data <- clean_dataframe(poi_data)
poi_data <- poi_data[!grepl("^Note|^Source|^Reference|^Please", poi_data$Residential.LGA.of.the.POI), ]

bocsar_variables <- c(
  "Table 2" = "Infringement notices",
  "Table 3" = "Young people proceeded against",
  "Table 4" = "Young people appearing in court",
  "Table 5" = "Young people in detention",
  "Table 6" = "Adults appearing in court",
  "Table 7" = "Adults in custody",
  "Table 8" = "Victims of violent crime"
)

process_bocsar_sheet <- function(df, sheet_name) {
  df %>%
    pivot_wider(names_from = Aboriginality, values_from = Count) %>%
    mutate(across(c(Aboriginal, `Non-Aboriginal`, Total), as.numeric)) %>%
    mutate(Year = case_when(
      sheet_name %in% c("Table 5", "Table 7") ~ as.numeric(sub("/.*", "", Year)),
      TRUE ~ as.numeric(as.character(Year))
    ),
    LGA = sapply(LGA, standardize_lga_name),
    BOCSAR_variable = if(sheet_name == "Table 1") Policing.Activity else bocsar_variables[sheet_name]) %>%
    dplyr::select(LGA, Year, Aboriginal, `Non-Aboriginal`, Total, BOCSAR_variable) %>%
    arrange(LGA, Year)
}

bocsar_processed <- map2(bocsar_data, names(bocsar_data), process_bocsar_sheet)

poi_processed <- poi_data %>%
  pivot_wider(names_from = Aboriginality, values_from = Value) %>%
  dplyr::select(-`NA`, -Unknown) %>%
  mutate(across(c(Aboriginal, `Non-Aboriginal`), as.numeric)) %>%
  mutate(Total = Aboriginal + `Non-Aboriginal`,
         LGA = sapply(Residential.LGA.of.the.POI, standardize_lga_name),
         Year = as.numeric(as.character(Year)),
         BOCSAR_variable = paste0("POI_", Offence)) %>%
  dplyr::select(LGA, Year, Aboriginal, `Non-Aboriginal`, Total, BOCSAR_variable) %>%
  filter(Year %in% c(2022, 2023)) %>%  # Filter POI data for 2022-2023
  arrange(LGA, Year)

bocsar_combined <- bind_rows(bocsar_processed)

# Combine all data, with POI data at the end
all_data <- bind_rows(bocsar_combined, poi_processed)




# Function to censor values
censor_values <- function(row) {
  censor_single <- function(x) {
    if (is.na(x) || x == "x") return("x")
    num_x <- suppressWarnings(as.numeric(x))
    if (is.na(num_x)) return(x)  # Keep non-numeric values as is
    ifelse(num_x < 5, "x", as.character(x))
  }
  
  censored <- lapply(row, censor_single)
  
  # Check if Aboriginal or Non-Aboriginal is "x" or if Total is less than 5
  aboriginal_censored <- censored$Aboriginal == "x"
  non_aboriginal_censored <- censored$`Non-Aboriginal` == "x"
  total_censored <- suppressWarnings(as.numeric(censored$Total)) < 5
  
  if (isTRUE(aboriginal_censored) || isTRUE(non_aboriginal_censored) || isTRUE(total_censored)) {
    censored$Total <- "x"
  }
  
  return(as.data.frame(censored))
}


# Apply censoring to the combined dataset
all_data_censored <- all_data %>%
  group_by(LGA, Year, BOCSAR_variable) %>%
  do(censor_values(.)) %>%
  ungroup()


# Load LGA shapes and join with censored data
LGA_shapes <- st_read("/Users/Current/Desktop/Good_Data_Institute/JustReinvest/data/LGA_2023_AUST_GDA2020")
LGA_shapes$LGA_NAME23_standardized <- sapply(LGA_shapes$LGA_NAME23, standardize_lga_name)
NSW_LGAs <- LGA_shapes %>% filter(STE_NAME21 == "New South Wales")
NSW_LGAs_simplified <- ms_simplify(NSW_LGAs, keep = 0.05, keep_shapes = TRUE)

bocsar_poi_spatial <- NSW_LGAs_simplified %>%
  left_join(all_data_censored, by = c("LGA_NAME23_standardized" = "LGA"))

bocsar_poi_spatial <- st_transform(bocsar_poi_spatial, 4326)

# Check for any remaining mismatches
data_mismatches <- setdiff(all_data_censored$LGA, NSW_LGAs$LGA_NAME23_standardized)
print("Remaining mismatches in combined data:")
print(data_mismatches)

save(bocsar_poi_spatial, file = "bocsar_poi_spatial.RData")
print(object.size(bocsar_poi_spatial), units = "Mb")
str(bocsar_poi_spatial)
```


```{r}
# Load required libraries
library(tidyverse)
library(readxl)
library(sf)
library(rmapshaper)

# ... (keep your existing code for loading BOCSAR data and processing it)

# Load population data
population_data <- read_excel("/Users/Current/Desktop/Good_Data_Institute/JustReinvest/code/GDI_BOCSAR_Dashboard/data/BOCSAR/BOCSAR_old/Aboriginal contacts with the Criminal Justice System by Local Government Area 2022-2023.xlsx", 
                              sheet = "Table 2", skip = 5)

# Clean and process population data
population_data <- population_data %>%
  select(LGA = LGA, 
         Aboriginal_Population = `Aboriginal census population`) %>%
  mutate(LGA = standardize_lga_name(LGA),
         Aboriginal_Population = as.numeric(Aboriginal_Population))

# Since the adult population is in Table 6, we need to load it separately
population_data_adult <- read_excel("/Users/Current/Desktop/Good_Data_Institute/JustReinvest/code/GDI_BOCSAR_Dashboard/data/BOCSAR/BOCSAR_old/Aboriginal contacts with the Criminal Justice System by Local Government Area 2022-2023.xlsx", 
                                    sheet = "Table 6", skip = 5)

# Clean and process adult population data
population_data_adult <- population_data_adult %>%
  select(LGA = LGA, 
         Aboriginal_Adult_Population = `Aboriginal census population aged 18+`) %>%
  mutate(LGA = standardize_lga_name(LGA),
         Aboriginal_Adult_Population = as.numeric(Aboriginal_Adult_Population))

# Merge population data
population_data <- population_data %>%
  left_join(population_data_adult, by = "LGA")

# Merge population data with bocsar_poi_spatial
bocsar_poi_spatial <- bocsar_poi_spatial %>%
  left_join(population_data, by = c("LGA_NAME23_standardized" = "LGA"))


# Read in  ROGS data for costs...

# Load required libraries
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)

# Load data
rogs_filepath <- "/Users/Current/Desktop/Good_Data_Institute/JustReinvest/code/GDI_BOCSAR_Dashboard/data/ROGS_combined/ROGS.xlsx"
Data8A <- read_excel(rogs_filepath, sheet = "8A")
Data17A <- read_excel(rogs_filepath, sheet = "17A")
Data6A <- read_excel(rogs_filepath, sheet = "6A")
Data7A <- read_excel(rogs_filepath, sheet = "7A")
DataCA <- read_excel(rogs_filepath, sheet = "CA")

# Extract data for Q1a1 (Police)
rogsq1a1 <- Data6A %>% 
  filter(Table_Number=="6A.1") %>% 
  filter(Description3=="Total recurrent expenditure") %>%
  mutate(across(NSW:NT, ~ as.numeric(gsub(",", "", .)))) %>%
  pivot_longer(cols = NSW:NT, names_to = "State", values_to = "Police") %>%
  select(Year, State, Police)
rogsq1a1$Police <- rogsq1a1$Police*1000

# Extract data for Q1a2 (Criminal Courts)
rogsq1a2 <- Data7A %>% 
  filter(Table_Number=="7A.11") %>% 
  filter(Description3=="All criminal courts") %>%
  mutate(across(NSW:NT, ~ as.numeric(gsub(",", "", .)))) %>%
  pivot_longer(cols = NSW:NT, names_to = "State", values_to = "Criminal") %>%
  select(Year, State, Criminal)

# Extract data for Q1a3 (Civil Courts)
rogsq1a3 <- Data7A %>% 
  filter(Table_Number=="7A.12") %>% 
  filter(str_detect(Court_Type,"All civil courts")) %>%
  mutate(across(NSW:NT, ~ as.numeric(gsub(",", "", .)))) %>%
  pivot_longer(cols = NSW:NT, names_to = "State", values_to = "Civil") %>%
  select(Year, State, Civil)

# Extract data for Q1a4 (Youth Justice)
rogsq1a4 <- Data17A %>% 
  filter(Table_Number=="17A.10") %>% 
  filter(Description3=="Detention-based services") %>%
  filter(Unit=="$'000") %>%
  mutate(across(NSW:NT, ~ as.numeric(gsub(",", "", .)))) %>%
  pivot_longer(cols = NSW:NT, names_to = "State", values_to = "Youth") %>%
  select(Year, State, Youth)

# Extract data for Q1a5 (Incarceration)
rogsq1a5 <- Data8A %>% 
  filter(Table_Number=="8A.1") %>% 
  filter(Description3=="Net operating expenditure") %>% 
  filter(Description4=="Total") %>%
  mutate(across(NSW:NT, ~ as.numeric(gsub(",", "", .)))) %>%
  pivot_longer(cols = NSW:NT, names_to = "State", values_to = "Incarceration") %>%
  select(Year, State, Incarceration)

# Merge data for Q1a
q1alist <- list(rogsq1a1, rogsq1a2, rogsq1a3, rogsq1a4, rogsq1a5) 
q1a <- Reduce(function(x,y) merge(x,y,all=TRUE), q1alist)
q1a$Year <- substr(q1a$Year, start=1, stop=4)
q1a <- pivot_longer(data=q1a, cols=Police:Incarceration, names_to="Sector", values_to="Expenditure")

# Extract data for Q1c2 (Criminal Courts cost per finalization)
rogsq1c2 <- Data7A %>% 
  filter(Table_Number=="7A.31") %>% 
  filter(Description3=="All criminal courts") %>%
  mutate(across(NSW:NT, ~ as.numeric(gsub(",", "", .)))) %>%
  pivot_longer(cols = NSW:NT, names_to = "State", values_to = "Criminal_Cost_Per_Finalization") %>%
  select(Year, State, Criminal_Cost_Per_Finalization)

# Extract data for Q1c5 (Youth Justice cost per day)
rogsq1c5 <- Data17A %>% 
  filter(Table_Number=="17A.21") %>% 
  filter(Description3=="Detention-based services") %>%
  filter(Description1=="Cost per average day per young person") %>%
  mutate(across(NSW:NT, ~ as.numeric(gsub(",", "", .)))) %>%
  pivot_longer(cols = NSW:NT, names_to = "State", values_to = "Youth_Cost_Per_Day") %>%
  select(Year, State, Youth_Cost_Per_Day)

# Extract data for Q1c7 (Incarceration cost per day)
rogsq1c7 <- Data8A %>% 
  filter(Table_Number=="8A.20") %>% 
  filter(Description2=="Per prisoner") %>% 
  filter(CS_Type=="Prison") %>%
  mutate(across(NSW:NT, ~ as.numeric(gsub(",", "", .)))) %>%
  pivot_longer(cols = NSW:NT, names_to = "State", values_to = "Incarceration_Cost_Per_Day") %>%
  select(Year, State, Incarceration_Cost_Per_Day)

# Merge data for Q1c
q1clist <- list(rogsq1c2, rogsq1c5, rogsq1c7) 
q1c <- Reduce(function(x,y) merge(x,y,all=TRUE), q1clist)
q1c$Year <- substr(q1c$Year, start=1, stop=4)

# Process ROGS data for NSW
rogs_processed <- q1a %>%
  filter(State == "NSW", Year == max(Year)) %>%
  select(Sector, Expenditure) %>%
  pivot_wider(names_from = Sector, values_from = Expenditure)

rogs_processed_per_person <- q1c %>%
  filter(State == "NSW", Year == max(Year)) %>%
  select(-State, -Year)

rogs_processed <- bind_cols(rogs_processed, rogs_processed_per_person)

# Print the processed ROGS data
print(rogs_processed)

# Add ROGS data to bocsar_poi_spatial (as state-wide constants)
bocsar_poi_spatial <- bocsar_poi_spatial %>%
  mutate(
    Police_Expenditure = rogs_processed$Police,
    Criminal_Courts_Expenditure = rogs_processed$Criminal,
    Civil_Courts_Expenditure = rogs_processed$Civil,
    Youth_Justice_Expenditure = rogs_processed$Youth,
    Incarceration_Expenditure = rogs_processed$Incarceration,
    Criminal_Courts_Cost_Per_Finalization = rogs_processed$Criminal_Cost_Per_Finalization,
    Youth_Justice_Cost_Per_Day = rogs_processed$Youth_Cost_Per_Day,
    Incarceration_Cost_Per_Day = rogs_processed$Incarceration_Cost_Per_Day
  )

# Save the updated data
save(bocsar_poi_spatial, file = "bocsar_poi_spatial_with_pop_and_costs.RData")
```


