---
title: "JustReinvest_EDA_Owen_V1"
format: html
editor: source
---

# Libraries

```{r}
library(tidyverse)
library(data.table)

library(ggplot2)
```



# Read in datasets
```{r}

rogs_6_police_df_raw <- fread("/Users/Current/Desktop/Good_Data_Institute/JustReinvest/data/drive-download-20240612T081926Z-001/rogs-2024-partc-section6-police-services-dataset.csv")

rogs_7_courts_df_raw <- fread("/Users/Current/Desktop/Good_Data_Institute/JustReinvest/data/drive-download-20240612T081926Z-001/rogs-2024-partc-section7-courts-dataset.csv")

rogs_8_corrective_df_raw <- fread("/Users/Current/Desktop/Good_Data_Institute/JustReinvest/data/drive-download-20240612T081926Z-001/rogs-2024-partc-section8-corrective-services-dataset.csv")

rogs_17_youth_df_raw <- fread("/Users/Current/Desktop/Good_Data_Institute/JustReinvest/data/drive-download-20240612T081926Z-001/rogs-2024-partf-section17-youth-justice-dataset_annotated.csv")

rogs_CA_overview_df_raw <- fread("/Users/Current/Desktop/Good_Data_Institute/JustReinvest/data/drive-download-20240612T081926Z-001/rogs-2024-partc-sector-overview-dataset.csv")


```



# EXPLORATORY ANALYSES / PLOTS

How much money is being spent on/in the criminal justice system, especially in disaggregated terms:

·  	By sector/service - police, courts, youth justice, incarceration, etc: CA.2; 7A.11 - total expenditure; 7A.31 - expenditure per finalisation

·   	On a per capita basis, by Aboriginality and age (e.g., per person per day, for youth and for adults)[of1] Possible derive variables from 8A.5 &  8A.7 - ATSI prisoner and community corrections population·      
 
How do ATSI spending and outcomes compare to the non-Indigenous population?

·   	Expenditures — possibly based on derived variables using 7A.11 and 8A.5/8A.7

·  	Incarceration rates -  possibly by 8A.5/8A.7 --> more info for Youth Justice 17A.18, 17A.28
 
 
 
 #-------
 
 
# Expenditure per finalisation
 

```{r}

# Filter for Table 7A.11 and exclude payroll tax
rogs_7_11 <- rogs_7_courts_df_raw %>% 
  filter(Table_Number == "7A.11") %>%
  filter(Description2 == "Excluding payroll tax") %>% 
  select(-c("Aust", "Aust cts"))

# Convert relevant columns to numeric
rogs_7_11 <- rogs_7_11 %>% 
  mutate(across(NSW:NT, ~ as.numeric(gsub(",", "", .))))

# Summarize the total expenditure by court type and year
total_expenditure_by_sector <- rogs_7_11 %>%
  group_by(Court_Type, Year) %>%
  summarise(across(NSW:NT, sum, na.rm = TRUE))

# Reshape the data for plotting
expenditure_long <- total_expenditure_by_sector %>%
  pivot_longer(cols = NSW:NT, names_to = "State", values_to = "Expenditure")


options(scipen = 9999)
# Plot the data
ggplot(expenditure_long, aes(x = Year, y = Expenditure, color = Court_Type, group = Court_Type)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") +
  theme_minimal() +
  labs(title = "Expenditure Over Time by Court Type and State/Territory",
       x = "Year",
       y = "Expenditure",
       color = "State/Territory") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


options(scipen = 0)

```




#--------------------


# Comparing Incarceration Rates

# Youth Justice

```{r}


# Filter for Table 7A.11 and exclude payroll tax
rogs_17A5 <- rogs_17_youth_df_raw %>% 
  filter(Table_Number == "17A.5",
         Service_Type == "Detention-based supervision",
         Age == "10-17 years old")


# Convert relevant columns to numeric
rogs_17A5 <- rogs_17A5 %>%
  mutate(across(NSW:Aust, ~ as.numeric(gsub(",", "", .))))

rogs_17A5_plotting_counts <- rogs_17A5 %>% filter(
  Description6!= "All people",
  Unit == "no."
) %>% 
  pivot_longer(cols = NSW:Aust, names_to = "State", values_to = "Number_Detention")

rogs_17A5_plotting_rates <- rogs_17A5 %>% filter(
  Description6!= "All people",
  Unit == "rate"
) %>% 
  pivot_longer(cols = NSW:Aust, names_to = "State", values_to = "Rate_Detention_per10000")


# Plot number of Indigenous youth in detention
ggplot(rogs_17A5_plotting_counts, aes(x = Year, y = Number_Detention, color = Description6, group = Description6)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") +
  theme_minimal() +
  labs(title = "Number of Youth in Detention Over Time",
       x = "Year",
       y = "Number of Youth",
       color = "State/Territory") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




# Plot Rates of Indigenous youth in detention
ggplot(rogs_17A5_plotting_rates, aes(x = Year, y = Rate_Detention_per10000, color = Description6, group = Description6)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") +
  theme_minimal() +
  labs(title = "Rates (per 10,000) of Youth in Detention Over Time",
       x = "Year",
       y = "Rates per 10,000",
       color = "State/Territory") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +ylim(0,65)


```



# Prison - adults

```{r}



# Filter for Table 8A.5
rogs_8A5 <- rogs_8_corrective_df_raw %>% 
  filter(Table_Number == "8A.5")


# Convert relevant columns to numeric
rogs_8A5 <- rogs_8A5 %>%
  mutate(across(NSW:Aust, ~ as.numeric(gsub(",", "", .))))

rogs_8A5_plotting_counts <- rogs_8A5 %>% filter(
  Indigenous_Status!= "All people",
  Unit == "no."
) %>% 
  pivot_longer(cols = NSW:Aust, names_to = "State", values_to = "Number_Prison")

rogs_8A5_plotting_rates <- rogs_8A5 %>% filter(
  Indigenous_Status!= "All people",
  Unit == "rate"
) %>% 
  pivot_longer(cols = NSW:Aust, names_to = "State", values_to = "Rate_Prison_per100000")


# Plot number of Indigenous people in detention
ggplot(rogs_8A5_plotting_counts, aes(x = Year, y = Number_Prison, color = Indigenous_Status, group = Indigenous_Status)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") +
  theme_minimal() +
  labs(title = "Number of People in Prison Over Time",
       x = "Year",
       y = "Number of People",
       color = "State/Territory") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




# Plot rates of Indigenous people in detention
ggplot(rogs_8A5_plotting_rates, aes(x = Year, y = Rate_Prison_per100000, color = Indigenous_Status, group = Indigenous_Status)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") +
  theme_minimal() +
  labs(title = "Rates (per 100,000) of People in Prison Over Time (Crude)",
       x = "Year",
       y = "Crude Rate per 100,000",
       color = "State/Territory") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +ylim(0,4200)



# Plot number of Indigenous youth in detention
ggplot(rogs_8A5_plotting_rates, aes(x = Year, y = Rate_Prison_per100000/10, color = Indigenous_Status, group = Indigenous_Status)) +
  geom_line() +
  facet_wrap(~ State, scales = "free_y") +
  theme_minimal() +
  labs(title = "Rates (per 10,000) of People in Prison Over Time (Crude)",
       x = "Year",
       y = "Crude Rate per 10,000",
       color = "State/Territory") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +ylim(0,420)


```

